\thetitle

### 2020-11-21 writing/challenges.tex
slight adjustment of challenges
Lots of minor changes and rewordings

\section{Why is it hard to evaluate blame?}

@@

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate the effect of these differences 
on its results, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems 
do not share a common baseline and, therefore, do not lend support to any
conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. This transient
variant accommodates varied linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

Beyond the implementation issue, the method also faces the problem that
the three variants---the natural semantics, the transient semantics, and
the erasure semantics---come with three radically different blame
assignment strategies.  For example, the natural semantics assigns blame
to one component while transient assigns blame to a set of components. The
erasure semantics does not blame a component per se, but it comes with an
exception location and the usual stack traces.

Supporting these strategies is {\em not\/} just an issue of language
implementation but affects greatly the behavior of the rational
programmer.  Specifically, the rational programmer must come with
\emph{modes} that represent the alternative interpretations of blame.  For
instance, a mode of the rational programmer using transient may pick the
oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode may opt to treat the blame
set as another form of a stack trace and add types to its newest element.
When both modes are equally successful in locating an impedance mismatch,
the method has needs to take into account the rational programmer's effort
to answer which mode is the most effective.

>>>>

\section{Why Is It Hard to Evaluate Blame?}
\label{sec:challenges}

@@

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate the effect of these differences 
on its results, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems 
do not share a common baseline and, therefore, do not lend support to any
conclusions about whether the value of blame changes from one system to
another. For instance, the case study presented here builds on
Greenman and Felleisen's Transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a Transient semantics that is
comparable with Typed Racket's default Natural one. This Transient
variant accommodates varied linguistic features and library calls
in a way that is faithful to the Transient
semantics~\cite{vss-popl-2017}.

Beyond the implementation issue, the method also faces the problem that
the three variants---the Natural semantics, the Transient semantics, and
the Erasure semantics---come with three radically different blame
assignment strategies.  For example, the Natural semantics assigns blame
to one component while Transient assigns blame to a set of components. The
Erasure semantics does not blame a component per se, but it comes with an
exception location and the usual stack traces.

Supporting these strategies is {\em not\/} just an issue of language
implementation but affects greatly the behavior of the rational
programmer.  Specifically, the rational programmer must come with
\emph{modes} that represent the alternative interpretations of blame.  For
instance, a mode of the rational programmer using Transient may pick the
oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode may opt to treat the blame
set as another form of a stack trace and add types to its newest element.
When both modes are equally successful in locating an impedance mismatch,
the method needs to take into account the rational programmer's effort
to answer which mode is the most effective.

### 2020-11-21 writing/conclusion.tex
Lots of minor changes and rewordings

%% -----------------------------------------------------------------------------
\section{What to Do Next}

The paper provides the first insight into the measurable value of blame
 in the gradually typed world. In an implicit manner researchers and
language designers have answered this question one way or another, without any
evidence. Now they may use the evaluation method of this paper to
confirm or reject their answers. We already know from~\citet{tgpk-dls-2018}
that, by intuition, programmers prefer the run-time checking of natural over
other soundness methods. But, just because a random set of programmers expresses
this view does not mean that blame assignment adds value. At this point we know
that it does, in some circumstances, and that in others, the answer remains
murky.

@@

The cited evidence suggests that this scenario is quite common and largely unadressed.
%, and that a type soundness check does {\em not\/} at all address the problem.  
Hence, a future evaluation must develop type-level mutators that produce incorrect type
annotations without breaking the code itself. Some preliminary work suggests
that such type mutators are much more difficult to develop than the code
mutators of this paper. We conjecture that the natural semantics will
usually discover such mistakes, while the transient and erasure semantics cannot
help with such mistakes at all; indeed, we expect that these latter two raise
misleading exceptions or produce plainly-wrong answers.  Of course, only further
evaluation work can confirm or reject these conjectures.

>>>>

%% -----------------------------------------------------------------------------
\section{What to Do Next} 
\label{sec:conclusion}

The paper provides the first insight into the measurable value of blame
 in the gradually typed world. In an implicit manner researchers and
language designers have answered this question one way or another, without any
evidence. Now they may use the evaluation method of this paper to
confirm or reject their answers. We already know from~\citet{tgpk-dls-2018}
that, by intuition, programmers prefer the run-time checking of Natural over
other soundness methods. But, just because a random set of programmers expresses
this view does not mean that blame assignment adds value. At this point we know
that it does, in some circumstances, and that in others, the answer remains
murky.

@@

The cited evidence suggests that this scenario is quite common and largely unadressed.
%, and that a type soundness check does {\em not\/} at all address the problem.  
Hence, a future evaluation must develop type-level mutators that produce incorrect type
annotations without breaking the code itself. Some preliminary work suggests
that such type mutators are much more difficult to develop than the code
mutators of this paper. We conjecture that the Natural semantics will
usually discover such mistakes, while the Transient and Erasure semantics cannot
help with such mistakes at all; indeed, we expect that these latter two raise
misleading exceptions or produce plainly-wrong answers.  Of course, only further
evaluation work can confirm or reject these conjectures.

### 2020-11-21 writing/discussion.tex
Lots of minor changes and rewordings

%% -----------------------------------------------------------------------------
\section{What We Learned}

%% \begin{itemize}
%% \item Why does the data look like this?
%%   - why tr fails
%%     - bad runtime errors
%%     - TR bug
%%   - why transient fails
%%     - bad runtime errors
%%     - timeouts/ooms: give context wrt fully untyped/typed versions
%%     - empty blame: trim Ben's prose

@@

The analysis in the last section indicates a number of interesting conclusions about blame in the gradual typing setting.
First, figure~\ref{fig:avo-matrix} makes it clear that any kind of dynamic checking has a large positive impact over Erasure, and this is true regardless of whether the rational programmer uses blame or stacktraces.
Indeed, the bottom row of figure~\ref{fig:avo-matrix} illustrates that Erasure almost never offers more useful information than any other mode.
Nonetheless, figure~\ref{fig:effort-table} does show that even if erasure often fails to provide helpful information, it provides the rational programmer with a quick answer.

@@

A second apparent flaw in our Transient semantics is the high cost of blame.
Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
worst-case of 5.4x on fully-typed benchmarks in Reticulated,
some of our configurations exceed a 4 minute timeout or 6GB memory limit with transient blame.
To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
Unfortunately, our high cost is closer to the truth;
 there are at least three broad issues that skew the earlier results.
First, the 2017 implementation of Reticulated fails to insert certain
 soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
 and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
 from the paper.
Second, Reticulated's type checker assigns the dynamic type to local
 variables in the "fully typed" benchmarks---often because the type system
 cannot articulate a more precise type.
Code that ends up with the dynamic type has fewer constraints to check at run-time.
Third, \citet{vss-popl-2017} use relatively small benchmarks.
Four have since been retired from the official Python benchmark suite
 because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
On larger programs, Reticulated suffers from high overhead.
For example, a Reticulated version of our \texttt{sieve} benchmark runs in
 ~40 seconds without blame, and with blame times out after 10 minutes.

>>>>

%% -----------------------------------------------------------------------------
\section{What We Learned}
\label{sec:discussion}

%% \begin{itemize}
%% \item Why does the data look like this?
%%   - why tr fails
%%     - bad runtime errors
%%     - TR bug
%%   - why Transient fails
%%     - bad runtime errors
%%     - timeouts/ooms: give context wrt fully untyped/typed versions
%%     - empty blame: trim Ben's prose

@@

The analysis in the last section indicates a number of interesting conclusions about blame in the gradual typing setting.
First, figure~\ref{fig:avo-matrix} makes it clear that any kind of dynamic checking has a large positive impact over Erasure, and this is true regardless of whether the rational programmer uses blame or stacktraces.
Indeed, the bottom row of figure~\ref{fig:avo-matrix} illustrates that Erasure almost never offers more useful information than any other mode.
Nonetheless, figure~\ref{fig:effort-table} does show that even if Erasure often fails to provide helpful information, it provides the rational programmer with a quick answer.

@@

A second apparent flaw in our Transient semantics is the high cost of blame.
Whereas \citet{vss-popl-2017} report an average slowdown of 2.5x and
worst-case of 5.4x on fully-typed benchmarks in Reticulated,
some of our configurations exceed a 4 minute timeout or 6GB memory limit with Transient blame.
To put those limits into context, the fully typed and fully untyped benchmarks all complete in a few seconds with minimal memory usage, and none of the mixed Natural configurations hit these limits.
Unfortunately, our high cost is closer to the truth;
 there are at least three broad issues that skew the earlier results.
First, the 2017 implementation of Reticulated fails to insert certain
 soundness checks\footnote{Missing check: \url{https://github.com/mvitousek/reticulated/issues/36}}
 and blame-map updates\footnote{Missing cast: \url{https://github.com/mvitousek/reticulated/issues/43}}
 from the paper.
Second, Reticulated's type checker assigns the dynamic type to local
 variables in the "fully typed" benchmarks---often because the type system
 cannot articulate a more precise type.
Code that ends up with the dynamic type has fewer constraints to check at run-time.
Third, \citet{vss-popl-2017} use relatively small benchmarks.
Four have since been retired from the official Python benchmark suite
 because they are too small, unrealistic, and unstable.\footnote{Release notes: \url{https://pyperformance.readthedocs.io/changelog.html}}
On larger programs, Reticulated suffers from high overhead.
For example, a Reticulated version of our \texttt{sieve} benchmark runs in
 ~40 seconds without blame, and with blame times out after 10 minutes.

### 2020-11-21 writing/ideas.tex
Lots of minor changes and rewordings

   For example we can obtain the simple blame mode
for the natural system if we use, as above, the selector $\blame{\cdot}$. 
However, the same mode has a different meaning in transient as this
semantics results in multiple blamed components.

@@

Of course for systems such as transient we can add modes where the rational
programmer picks one blametrail as part of the migration process. For
example the \first{\blame{\cdot}} mode continues with the oldest blame
entry in the blameset of a configuration, as we discuss in
section~\ref{sec:challenges}.

>>>>

   For example we can obtain the simple blame mode
for the Natural system if we use, as above, the selector $\blame{\cdot}$. 
However, the same mode has a different meaning in Transient as this
semantics results in multiple blamed components.

@@

Of course for systems such as Transient we can add modes where the rational
programmer picks one blametrail as part of the migration process. For
example the \first{\blame{\cdot}} mode continues with the oldest blame
entry in the blameset of a configuration, as we discuss in
section~\ref{sec:challenges}.

### 2020-11-21 writing/introduction.tex
Lots of minor changes and rewordings

\section{Does Blame Matter?}

@@

This paper's first contribution is an automated method for evaluating the
effectiveness of blame assignment strategies in the gradual typing world.
%% (sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
%% MF: says you can't simulate a roadmap by listing all sections.
%% MF: and for heaven's sake lookup the use of ~ in TeX
Roughly speaking, the method, which is
loosely based on the work of Lazarek et al.~\cite{lksfd-popl-2020},
simulates a programmer who follows the Wadler-Findler slogan quoted above.
Such a \emph{rational programmer} can find impedance-mismatch
bugs by adding types to a partially typed program (section~\ref{sec:landscape}). That is, when the
run-time system signals an impedance mismatch, a good blame message helps
the programmer infer where to add an additional type specification. Since
components that pass the type checker cannot cause impedance mismatches;
the type checker discovers these mismatches at compile time and points
directly to the faulty component. Otherwise, if the
additional type does {\em not\/} cause a static type error, the program
 is run again to signal another violation. Then the programmer can heed
blame one more time to add yet another type specification. Weighing the number of
steps it takes to discover the mistake in this manner is a measure of the effectiveness
of the blame assignment strategy of the language.

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment strategies and approximately
72000 cases (section~\ref{sec:results}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language. A direct interpretation of the results
(section~\ref{sec:discussion}) is that
they validate the conjecture behind the work of theoreticians. A good blame assignment
strategy helps the search for impedance mismatches between the specified types
and the behavior of untyped components. Even better, the evaluation method 
identifies the wrapper-based blame tracking of Typed Racket as the most
useful academic system with that of Reticulated Python only marginally  behind.

>>>>

\section{Does Blame Matter?}
\label{sec:introduction}

@@

This paper's first contribution is an automated method for evaluating the
effectiveness of blame assignment strategies in the gradual typing world.
%% (sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
%% MF: says you can't simulate a roadmap by listing all sections.
%% MF: and for heaven's sake lookup the use of ~ in TeX
Roughly speaking, the method, which is
loosely based on the work of Lazarek et al.~\cite{lksfd-popl-2020},
simulates a programmer who follows the Wadler-Findler slogan quoted above.
Such a \emph{rational programmer} can find impedance-mismatch
bugs by adding types to a partially typed program (section~\ref{sec:landscape}). That is, when the
run-time system signals an impedance mismatch, a good blame message helps
the programmer infer where to add an additional type specification.
Components that pass the type checker cannot cause impedance mismatches,
so if the type checker discovers these mismatches at compile time then it points
directly to the faulty component. Otherwise, if the
additional type does {\em not\/} cause a static type error, the program
is run again to signal another violation. Then the programmer can heed
blame one more time to add yet another type specification.
Weighing whether we discover the mistake in this manner, and the number of
steps it takes to do so, are measures of the effectiveness
of the blame assignment strategy of the language.

The paper's second contribution is a set of results from applying the evaluation
method to three distinct blame assignment strategies and approximately
722,000 cases (section~\ref{sec:results}): Vitousek et
al.'s strategy of tracking typed/untyped boundaries in Reticulated;
Tobin-Hochstadt et al.'s use of higher-order contract system~\cite{ff-icfp-2002,
mf-toplas-2009}; and the approach of industrial systems that forego blame
and instead rely on error messages from the
underlying programming language. A direct interpretation of the results
(section~\ref{sec:discussion}) is that
they validate the conjecture behind the work of theoreticians. A good blame assignment
strategy helps the search for impedance mismatches between the specified types
and the behavior of untyped components. Even better, the evaluation method 
identifies the wrapper-based blame tracking of Typed Racket as the most
useful academic system with that of Reticulated Python only marginally  behind.

### 2020-11-21 writing/landscape.tex
Lots of minor changes and rewordings

\section{How to Bring Natural, Transient and Erasure Under the Same Roof}

@@

The  program in figure~\ref{fig:example} demonstrates the differences
between natural, transient and erasure using Typed Racket syntax.  It
consists of four modules: the left and right ones are untyped while
the two modules in the middle are typed.

@@

Under natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lb} or
\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the run time wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of its type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the natural semantics blames \texttt{pack-lib}.

Under the transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
produce an error. In contrast to natural, transient does not rely on
proxies. Instead, it rewrites typed modules to inline checks at appropriate
places. Furthermore, these checks are
partial, that is, they check only the top constructor of a type. In other words
they check types in a  \emph{shallow} manner. In our
example, transient's inlined checks make sure that the imported \texttt{pack} 
is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
order to pinpoint the responsible party for the failed checked, transient
keeps on the side a map from values to the boundaries between modules they
cross and the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to
\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
\texttt{client} with the type that appears in the \texttt{required/typed}
forms in the example. Since the failed check corresponds to the return
type of \texttt{pack}, transient blames the sources of the two boundary
crosses, which in this case is \texttt{untyped/pack-lib} for both. In
general though, transient may blame more than one parties. In fact, the
theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
where transient constructs a blame set that includes modules that are not
relevant to the check failure or that it even misses entirely the responsible
party ---  our experiment aims to examine the pragmatic impact of exactly these
issues on debugging type-level mistakes.

Under erasure semantics, the example does not raise an error due to the
gradual type system; erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce an wrong result that goes undetected by the
gradual type system .  In either case, the erasure semantics provides no
help to the programmer for detecting the bug in our program.

Since, the goal of our experiment is to compare the effectiveness of blame
across the three semantics, it is critical to isolate their
distinguishing characteristics from superficial implementation
differences. Thus we implement all three systems on top of the common
basis of Racket. First, we use the Typed Racket syntax and type
checker for all three semantics. Second, for natural we use Typed Racket as
is while for erasure we run programs with plain Racket after erasing any
type annotations. Third, for transient we implement a variant of Typed
Racket that follows closely the semantics from~\citet{vss-popl-2017}.

The first major ingredient of transient Typed Racket is to determine the
Racket implementation of the shallow checks that map to each type.  For
certain types, such as {\tt String} or {\tt (Vectorof Byte)}, transient
can re-use a predicate from the Racket language.  Other types require custom
predicates; for example the type {\tt Index} of valid array indices allows
a range of positive exact integers.  Many custom predicates run in
constant time, but some are more involved.  One linear-time check is for
{\tt (Listof T)} for any element type {\tt T}, because it must distinguish
proper from improper list-like structures.  The
key point, however, is that no shallow check inspects beyond the top
constructor of each type.

The second major ingredient of transient Typed Racket  is the addition of
a pass to the compiler of Typed Racket that rewrites all typed code.
Rewriting modifies all elimination forms to shallow-check their result.
For example, if the variable {\tt f} has the static type {\tt (-> String
String)} then the application {\tt (f x)} must be rewritten to check for a
string result because {\tt f} could be an untyped function.  Rewriting
also inserts hooks to update the blame map.  When a value crosses a module
boundary, the blame map records a type assumption.  The inlined checks
for elimination forms also extend the blame map with further information. For
instance, the check of the result of {\tt (f x)} adds to the blame map
that the expected type of the result is the return type of the type of
{\tt f}.  If a transient check ever fails, transient Typed Racket walks over 
the information in the blame map and reports a set of blamed components.

>>>>

\section{How to Bring Natural, Transient and Erasure Under the Same Roof} 
\label{sec:landscape}

@@

The  program in figure~\ref{fig:example} demonstrates the differences
between Natural, Transient and Erasure using Typed Racket syntax.  It
consists of four modules: the left and right ones are untyped while
the two modules in the middle are typed.

@@

Under Natural semantics~\cite{tf-popl-2008,tfffgksst-snapl-2017}, when
\texttt{pack} is imported in either \texttt{typed/pack-lb} or
\texttt{typed/secure-pack-lb}, the runtime checks that \texttt{pack} is a
function and wraps it in a proxy that enforces its  type.  Analogously
the run time wraps the exports of the two typed modules in proxies that
check that untyped code applies \texttt{typed-pack} and
\texttt{crypto-pack} to arguments of its type. In essence, the
return type checks protect  the typed modules in case they use
\texttt{pack} and, as is the case here, it does not return an association
list. The check of the argument type  treats the
exports of the two typed modules as typed values that need to be protected
when they are used by untyped code.  As this analysis implies, if the
return type check fails then the problem is with
\texttt{untyped/pack-lib}, while if the argument type check fails
responsibility lies with \texttt{client}. The Natural semantics keep track
of the responsible party for each check in the corresponding proxy. Thus,
in our example, when the call to \texttt{typed-pack} returns, the return
type check fails and the Natural semantics blames \texttt{pack-lib}.

Under the Transient semantics~\cite{vss-popl-2017}, the call to \texttt{typed-pack} does not
produce an error. In contrast to Natural, Transient does not rely on
proxies. Instead, it rewrites typed modules to inline checks at appropriate
places. Furthermore, these checks are
partial, that is, they check only the top constructor of a type. In other words
they check types in a  \emph{shallow} manner. In our
example, Transient's inlined checks make sure that the imported \texttt{pack} 
is a function and that the result of its call in \texttt{typed/crypto-pack-lib} is a list.   Exactly this
last check fails when \texttt{client} calls to \texttt{crypto-pack}. In
order to pinpoint the responsible party for the failed checked, Transient
keeps on the side a map from values to the boundaries between modules they
cross and the corresponding types. In the example, the map records that
\texttt{pack} crosses from \texttt{untyped/pack-lib} to
\texttt{typed/pack-lib} and from \texttt{typed/pack-lib} to
\texttt{client} with the type that appears in the \texttt{required/typed}
forms in the example. Since the failed check corresponds to the return
type of \texttt{pack}, Transient blames the sources of the two boundary
crosses, which in this case is \texttt{untyped/pack-lib} for both. In
general though, Transient may blame more than one parties. In fact, the
theoretical work of~\citet{gfd-oopsla-2019} shows that there are programs
where Transient constructs a blame set that includes modules that are not
relevant to the check failure or that it even misses entirely the responsible
party ---  our experiment aims to examine the pragmatic impact of exactly these
issues on debugging type-level mistakes.

Under Erasure semantics, the example does not raise an error due to the
gradual type system; Erasure performs no type checks at runtime. Thus the
program may fail with an exception if \texttt{client} tries to use the
result of \texttt{typed-pack} or \texttt{crypto-pack}, but it may also not
fail at all and produce an wrong result that goes undetected by the
gradual type system .  In either case, the Erasure semantics provides no
help to the programmer for detecting the bug in our program.

Since, the goal of our experiment is to compare the effectiveness of blame
across the three semantics, it is critical to isolate their
distinguishing characteristics from superficial implementation
differences. Thus we implement all three systems on top of the common
basis of Racket. First, we use the Typed Racket syntax and type
checker for all three semantics. Second, for Natural we use Typed Racket as
is while for Erasure we run programs with plain Racket after erasing any
type annotations. Third, for Transient we implement a variant of Typed
Racket that follows closely the semantics from~\citet{vss-popl-2017}.

The first major ingredient of Transient Typed Racket is to determine the
Racket implementation of the shallow checks that map to each type.  For
certain types, such as {\tt String} or {\tt (Vectorof Byte)}, Transient
can re-use a predicate from the Racket language.  Other types require custom
predicates; for example the type {\tt Index} of valid array indices allows
a range of positive exact integers.  Many custom predicates run in
constant time, but some are more involved.  One linear-time check is for
{\tt (Listof T)} for any element type {\tt T}, because it must distinguish
proper from improper list-like structures.  The
key point, however, is that no shallow check inspects beyond the top
constructor of each type.

The second major ingredient of Transient Typed Racket  is the addition of
a pass to the compiler of Typed Racket that rewrites all typed code.
Rewriting modifies all elimination forms to shallow-check their result.
For example, if the variable {\tt f} has the static type {\tt (-> String
String)} then the application {\tt (f x)} must be rewritten to check for a
string result because {\tt f} could be an untyped function.  Rewriting
also inserts hooks to update the blame map.  When a value crosses a module
boundary, the blame map records a type assumption.  The inlined checks
for elimination forms also extend the blame map with further information. For
instance, the check of the result of {\tt (f x)} adds to the blame map
that the expected type of the result is the return type of the type of
{\tt f}.  If a Transient check ever fails, Transient Typed Racket walks over 
the information in the blame map and reports a set of blamed components.

### 2020-11-21 writing/mutate.tex
Put figure 2 in one column
Lots of minor changes and rewordings

\section{How to Make Lots of Mistakes}

@@

\begin{figure*}
\begin{tabular}{p{2cm} | p{10cm} }
  {\bf  name} & {\bf description (author)}  \\

@@

\end{tabular}
  \caption{Benchmarks summary.}
  \label{table:benchmark-descriptions}
\end{figure*}

The starting point for our corpus of programs
is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
benchmark suite consists of fully typed correct programs that are written by different authors
and have been used, maintained and evolved by their authors and others over a
number of years.  They range widely in size, complexity, purpose and
features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
of Typed Racket's type system such as occurrence
typing~\cite{tf-icfp-2010}, types for mutable and immutable data
structures, polymorphic types, types for first-class classes and objects and types for
Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
we select the ten largest in terms of numbers of components (between 6 and 14 components).
These benchmarks also have the most complex dependency graphs and, hence,
can make the debugging process the hardest for the rational programmer.
The table in figure~\ref{table:benchmark-descriptions} shows our selection
together with a short description for each benchmark.

@@

The astute reader must have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{scenarios}, of the benchmarks without 
some  of their type annotation.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

@@

With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
are for turning the benchmarks into interesting debugging scenarios.
Indeed the mutators produce at least 72192 interesting debugging scenarios
across all mutants --- there are potentially many more but this is how
many we examined. In more detail, the mutators produce at least 40 interesting
scenarios for every benchmark, and these scenarios originate from at
least four different mutators per benchmark.  Thus the mutators result in 
a sizeable and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one
benchmark. Especially some mutators  target specific features of
benchmarks even if they do not contribute interesting scenarios to other
benchmarks, which
justifies their inclusion. For
instance the class-focused mutators are effective in \texttt{take5}
which jives with the fact that it is the benchmark with 
the most extensive use of object-oriented features.

>>>>

\section{How to Make Lots of Mistakes} 
\label{sec:mutate}

@@

\begin{figure}
\begin{tabular}{p{1.75cm} | p{6cm} }
  {\bf  name} & {\bf description (author)}  \\

@@

\end{tabular}
  \caption{Benchmarks summary.}
  \label{table:benchmark-descriptions}
\end{figure}

The starting point for our corpus of programs
is~\citet{gtnffvf-jfp-2019}'s gradual typing benchmark suite. The
benchmark suite consists of fully typed correct programs that are written by different authors
and have been used, maintained and evolved by their authors and others over a
number of years.  They range widely in the size, complexity, purpose and
features of Typed Racket they employ.  Furthermore the benchmarks use advanced aspects
of Typed Racket's type system such as occurrence
typing~\cite{tf-icfp-2010}, types for mutable and immutable data
structures, polymorphic types, types for first-class classes and objects, and types for
Racket's numeric tower~\cite{stathff-padl-12}. Without loss of the diversity of the benchmarks,
we select the ten largest in terms of numbers of components (between 6 and 14 components).
These benchmarks also have the most complex dependency graphs and, hence,
can make the debugging process the hardest for the rational programmer.
The table in figure~\ref{table:benchmark-descriptions} shows our selection
together with a short description for each benchmark.

@@

The astute reader must have observed that mutators that produce 
type-level mistakes do not necessarily turn our benchmarks into
interesting debugging scenarios for the rational programmer.  The
benchmarks are fully typed, and type checking detects each mistake at
compile time.  Hence, we apply the mutators to versions, dubbed
\dubbed{scenarios}, of the benchmarks without 
some  of their type annotations.  Without the annotations, the type
checker may not detect a mistake at compile time and, instead, the run
time checks
of the gradual type system signal an error.  Scenarios
with type-level mistakes that result in run time errors are the ideal
ground to exercise the rational programmer's ability to locate errors.

@@

With the definition of interesting debugging scenarios in hand, we analyze how effective our mutators
are for turning the benchmarks into interesting debugging scenarios.
At a high level, the mutators produce 16,800 mutants with at least one interesting debugging scenario each across all of our benchmarks.
Broken down by benchmark, the mutators produce at least 40 interesting
scenarios for every benchmark, and these scenarios originate from at
least four different mutators per benchmark.  Thus the mutators result in 
a sizeable and diverse population of scenarios for every benchmark.
Furthermore, every mutator contributes scenarios to at least one
benchmark.
That being said, some mutators only apply to a small subset of the benchmarks;
those mutators target specific features which only those benchmarks make us of, so we include them despite their narrow applicability.
For instance, the class-focused mutators are mainly effective in \texttt{take5}, since it is the benchmark with the most extensive use of object-oriented features.
